Finite Complete Suites for CSP Refinement Testing

### 2018-11-16 scp-journal/article/preliminaries.tex
Re-formatted for Science of Computing journal (new directory scp-journal). Re-worked discussion.
revised sections 1 and 2, shortenned paper

We present CSP~(Section~\ref{section:csp}) and the concept of minimal hitting
sets~(Section~\ref{sec:hit}), which is central to our novel notion of test
for failures refinement. To study complexity, we also introduce the concept
of Sperner families~(Section~\ref{sec:sperner}).

@@

\subsubsection*{Communicating Sequential Processes (CSP)} is a process algebra
supporting system development by refinement. Using CSP, we model both systems and
their components using processes. They are characterised by their patterns
of interactions, modelled by synchronous, instantaneous, and atomic events.

Throughout this paper, the alphabet of the processes, that is, the set of
events that are in scope, is denoted by $\Sigma$ and supposed to be finite.
The FDR tool~\cite{fdr} supports model checking and semantic analyses of
finite-state CSP processes.

@@

Two forms of choice support branching behaviour. An external choice $P
\extchoice Q$ between processes $P$ and $Q$ offers to the environment the
initial events of $P$ and $Q$. Once synchronisation takes place, the process
that has offered this event is chosen and the other is discarded. In an
internal choice $P \intchoice Q$, the environment does not have an
opportunity to interfere:~the choice is made by the process.
%
\begin{example}\label{example:CSP}
  We consider $P$, $Q$, and $R$ below. $P$ is initially
  ready to engage in the event $a$, and then makes an internal  choice to
  behave like either $Q$ or $R$.
  \begin{eqnarray*}
  P & = & a \then (Q\intchoice R)
  \\
  Q & = & a \then P \extchoice c \then P
  \\
  R & = & b \then P \extchoice c \then R
  \end{eqnarray*}
  $Q$, for instance, offers to the environment the choice to engage in $a$
  again or $c$. In both cases, afterwards, we have a recursion back to $P$.
  In $R$, if $b$ is chosen, we also have a recursion back to $P$. If $c$ is
  chosen, the recursion is to $R$.
  \xbox
\end{example}
%
Iterated forms $\Extchoice i: I @ P(i)$ and $\Intchoice i: I @ P(i)$ of the
external and internal choice operators define a choice over a collection of
processes $P(i)$. If the index set $I$ is empty, the external choice is the
process $\Stop$, which deadlocks:~does not engage into any event or
terminate. For an external choice $\Extchoice e: A @ e \then P(e), A\subseteq
\Sigma$ over a set $A$ of events, we use the abbreviation $e:A \then P(e)$.
An iterated internal choice is not defined for an empty index set.

There are several parallelism operators. A widely used form of parallelism $P
\parallel[cs] Q$ defines a process in which the behaviour is characterised by
those of $P$ and $Q$ in parallel, synchronising on the events in the set
$cs$. Other forms of parallelism available in CSP can be defined using this
operator.

@@

Semantic definitions specify, for each operator, how the traces or failures
of the resulting process can be calculated from the traces or failures of
each operand. For example, for internal choice, $\fails(P\intchoice Q) =
\fails(P)\cup\fails(Q)$; see~\cite[p.~210]{Roscoe:1997:TPC:550448} for a
comprehensive list of these definitions.

Using the notation $P/s$ to denote the process $P$ after having engaged into
trace $s$, the set $\refs(P/s) \defs \{~X | (s,X) \in \fails(P)~\}$ contains
the  refusals of $P$ after $s$. Refusals are
subset-closed~\cite{Hoare:1985:CSP:3921,Roscoe2010}: if $(s,X)$ is a failure
of $P$ and $Y\subseteq X$, then $(s,Y)\in\fails(P)$ and $Y\in\refs(P/s)$
follows.

For divergence-free processes, failures refinement, $P \lessdet_F Q$, is
defined as expected by set containment $\fails(Q)\subseteq\fails(P)$. Since
refusals are subset-closed, $P \lessdet_F Q$ implies
$(s,\varnothing)\in\fails(P)$ for all traces $s\in\trc(Q)$. So, for
divergences-free processes, failures refinement implies trace refinement.
Therefore, using the conformance relation $conf$ below
%
\begin{equation}\label{eq:conf}
  Q\ conf\ P \defs \forall s\in \trc(P) \cap \trc(Q): \refs(Q/s)
  \subseteq \refs(P/s),
\end{equation}
%
failures refinement can be expressed by $\lessdet_T$ and $conf$.
%
\begin{equation}\label{thm:fref-tref-conf}
(P \lessdet_F Q) \iff (P \lessdet_T Q \land Q\ conf\ P)
\end{equation}
%
For finite CSP processes, since refusals are
subset-closed, $\refs(P/s)$ can be constructed from the set of \emph{maximal
refusals} defined by
%
\begin{equation}
\maxrefs(P/s) = \{ R \in\refs(P/s)~|~\forall R'\in \refs(P/s) - \{ R\}: R \not\subseteq R'\}
\end{equation}
%
Conversely, with the maximal refusals $\maxrefs(P/s)$ at hand, we can
reconstruct the refusals in the set $\refs(P/s)$ as described by
%
\begin{equation}\label{eq:refMaxRef}
\refs(P/s) = \{ R'\in 2^\Sigma~|~\exists R\in \maxrefs(P/s): R'\subseteq R \}.
\end{equation}
%
Deterministic process states $P/s$ have exactly the one maximal refusal
defined by $\Sigma-[P/s]^0$, where $[P/s]^0$ denotes the \emph{initials} of
$P/s$, that is, the events that $P/s$ may engage into. The maximal refusals
in combination with the initials of a process express its degree of
nondeterminism.
%
\begin{example}
\label{ex:nondetdegree} $P = (\Stop \intchoice Q)$ has maximal refusals
$\maxrefs(P) = \{ \Sigma \}$, because $\Stop$ refuses to engage in any event,
and this is carried over to $P$ by the internal choice. However, $P$ is
distinguished from $\Stop$ by its initials, which are defined by $[P]^0 = [
\Stop \intchoice Q]^0 = [Q]^0$. So $P$ may engage nondeterministically in any
initial event of $Q$, but also refuse everything, due to internal selection
of $\Stop$.

Assuming an alphabet $\Sigma = \{a,b,c,d\}$, the process
%
$$Q = (e:\{a,b\} \then \Stop) \intchoice (e:\{c,d\} \then \Stop)$$
%
has maximal refusals $\maxrefs(Q) = \{ \{c,d\}, \{a,b\} \}$ and initials
$[Q]^0=\Sigma$. In contrast to $P$, nondeterminism is reflected here by
two maximal refusals. \xbox
\end{example}

% =========================================================================
\subsubsection*{Normalised Transition Graphs for CSP Processes}
\label{sec:ntg}

As shown in~\cite{Roscoe:1994:chapter}, the failures semantics of any
finite-state CSP process $P$ can be represented by a \emph{normalised
transition graph} $G(P)$ defined by a tuple
$$
G(P) = ( N, \ii n, \Sigma, t : N\times\Sigma \pfun N, r : N \fun \mathbb{P}\mathbb{P}(\Sigma)),
$$
with nodes $N$, initial node $\ii n\in N$, and process alphabet $\Sigma$. The
partial \emph{transition function} $t$ maps a node $n$ and an event
$e\in\Sigma$ to its successor node $t(n,e)$. If $(n,e)$ is in the domain of
$t$, then there is a transition, that is, an outgoing edge, from $n$ with
label $e$, leading to node $t(n,e)$. Normalisation of $G(P)$ is reflected by
the fact that $t$ is a function. The graph construction
in~\cite{Roscoe:1994:chapter} implies that all nodes $n$ in $N$ are
reachable by sequences of edges labelled by $e_1\dots e_k$ and connecting
states $\ii n,n_1,\dots,n_{k-1},n$, such that
\[
n_1 = t(\ii n,e_1), \quad n_i = t(n_{i-1},e_i),\ i = 2,\dots,k-1,\quad
n= t(n_{k-1},e_k).
\]
%
By construction, $s\in\Sigma^*$ is a trace of $P$, if, and only if, there is
a path through $G(P)$ starting  at $\ii n$ whose edge labels coincide with
$s$. In analogy to $\trc(P)$, we use the notation $\trc(G(P))$ for the set of
finite, initialised paths through $G(P)$, each path represented by its finite
sequence of edge labels. We note that $\trc(P) = \trc(G(P))$. Since $G(P)$ is
normalised, there is a unique node reached by following the events from $s$
one by one, starting in $\ii n$. Therefore, $G(P)/s$  is also well defined.
By $[n]^0$ we denote the \emph{initials} of $n$:~the set of events occurring
as labels in any outgoing transitions.
$$
[n]^0 = \{ e\in\Sigma~|~(n,e)\in\dom~t \}
$$
The graph construction from~\cite{Roscoe:1994:chapter} used to define
$G(P)$ for any process $P$ guarantees that $[G(P)/s]^0 = [P/s]^0$ for all
traces $s$ of $P$.

@@

% =========================================================================
\subsubsection*{Acceptances}
\label{sec:accs}

When investigating  tests for failures refinement, the notion of
\emph{acceptances}, which is dual to refusals, is also useful. While the
original introduction of acceptances presented
in~\cite[pp.~75]{Hennessy:1988:ATP:50497} was independent of refusals, we use
the definition from~\cite[pp.~278]{Roscoe:1997:TPC:550448}.   A \emph{minimal
acceptance} of a CSP process state $P/s$ is the complement of a maximal
refusal of the same state. The set of minimal acceptances of $P/s$ is denoted
by $\minaccs(P/s)$ and formally defined as
%
\begin{equation}\label{eq:accref}
\minaccs(P/s) = \{ \Sigma - R~|~R\in \maxrefs(P/s)  \}
\end{equation}
%
With this definition, a (not necessarily minimal) \emph{acceptance} of $P/s$ is a superset of some minimal acceptance and a subset of the initials $[P/s]^0$. Denoting the acceptances of $P/s$ by $\accs(P/s)$, this leads to the formal definition
\begin{equation}
\label{eq:accrefall}
\accs(P/s)  = \{ B\subseteq [P/s]^0~|~\exists A\in\minaccs(P/s):  A \subseteq B   \}
\end{equation}
%
Acceptances have the following intuitive interpretation.  If the behaviour of
$P/s$ is deterministic, its only acceptance equals $[P/s]^0$, because $P/s$
never refuses any of the events in this set. If $P/s$ is nondeterministic, it
internally chooses one of its \emph{minimal acceptance sets} $A$ and never
refuses any event in $A$, while {\it possibly} refusing the events from
$[P/s]^0 - A$ and {\it always} refusing those from $\Sigma - [P/s]^0$.

Exploiting (\ref{eq:accref}),
the nodes of a normalised transition graph can
alternatively be labelled with their minimal acceptances, and this
information is equivalent to that contained in the maximal refusals. Since
process states $P/s$ are equivalently expressed by states $G(P)/s$ of $P$'s
normalised transition graph, we also write $\minaccs(G(P)/s)$ and
note that (\ref{eq:refsAndR}) and (\ref{eq:accref}) imply
\begin{equation}\label{eq:maxrefsminaccs}
\minaccs(G(P)/s) = \{ \Sigma - R | R\in r(G(P)/s) \} = \minaccs(P/s).
\end{equation}
%
Given any non-diverging, non-terminating, finite-state process $P$, it can be
re-con\-struc\-ted from its graph $G(P)$ with initial state $\ii
n$ and transition function $t$, using $P$'s normalised syntactic
representation~\cite[pp.~277]{Roscoe:1997:TPC:550448} specified as follows.
\begin{eqnarray*}
\text{normalised}(P) & = & P_N(\ii n)
\\
P_N(n) & = & \Intchoice_{A\in\minaccs(n)\cup \{ [n]^0 \} } e:A\then P_N(t(n,e))
\end{eqnarray*}
With this definition, it is established that $P$ is semantically equivalent
to the process $\text{normalised}(P)$ in the CSP failures semantics.

% .....................................................................................
 \begin{figure}[t]
   %%\hspace*{-40mm}
   \begin{center}
     \includegraphics[width=\textwidth]{p.pdf}
   \end{center}
   %%\vspace*{-10mm}
   \caption{Normalised transition graph of CSP process $P$ from Example~\ref{ex:a}.}
   \label{fig:tga}
 \end{figure}

% .......................................................................................

@@

node~3 corresponds to process state $R/c = R$, and therefore it is labelled
by $R$'s minimal acceptance $\{b,c\}$. \xbox

@@

Summarising, refinement relations between finite-state CSP processes $P, Q$ can be be
expressed by means of their normalised transition graphs

@@

  \label{eq:failconf}
  P \lessdet_F Q & \Leftrightarrow & P \lessdet_T Q \wedge Q\ conf\ P

@@

Recall that $P\lessdet_T Q$ is defined as $\trc(Q)\subseteq \trc(P)$ and that
$\trc(P)=\trc(G(P))$ and $\trc(Q)=\trc(G(Q))$. This shows (\ref{eq:trcrefa}).
Result (\ref{eq:failconf}) has been shown
in~\cite{DBLP:conf/icfem/CavalcantiG07}.

@@

\\\tabc

@@

@@

% =========================================================================
\subsubsection*{Reachability Under Sets of Traces}
\label{sec:V} Given a finite-state CSP process $P$ and its normalised
transition graph $G(P)$,
%\[
%G(P) = ( N, \ii n, \Sigma, t : N\times\Sigma \pfun N, r : N \fun \mathbb{P}\mathbb{P}(\Sigma)),
%\]
we suppose that $V\subseteq\Sigma^*$ is a prefix-closed set  of sequences of
events. By $t(\ii n,V)$ we denote the set
\[
t(\ii n,V) = \{ n\in N~|~\exists s\in V: s\in\trc(P)\wedge G(P)/s = n \}
\]
of nodes in $N$ that are reachable in $G(P)$ by applying traces of $V$. The
lemma below specifies a construction method for such sets $V$ reaching {\it
every} node of $N$.

\begin{lemma}
\label{lemma:extendV} Let $P$ be a CSP process with normalised transition
graph $G(P)$. Let $V\subseteq\Sigma^*$ be a finite prefix-closed set of
sequences of events. Suppose that  $G(P)$ reaches $k < |N|$ nodes under $V$,
that is, $|t(\ii n,V)| = k$. Let $V.\Sigma$ denote the set of all sequences
from $V$, extended by any event of $\Sigma$. We can then conclude that $G(P)$
reaches at least $(k+1)$ nodes under $V\cup V.\Sigma$.
\end{lemma}
\begin{proof}
Suppose that $n'\in (N - t(\ii n,V))$.  Since all nodes in $N$ are reachable,
there exists a trace $s$ such that $G(P)/s = n'$. Decompose $s = s_1.e.s_2$
with $s_i\in\Sigma^*, e\in\Sigma$, such that $G(P)/s_1 \in t(\ii n,V)$ and
$G(P)/s_1.e \not\in t(\ii n,V)$. Such a decomposition always exists, because
$V$ is prefix-closed and therefore contains the empty trace $\varepsilon$.
Note, however, that it is not necessarily the case that $s_1\in V$.

Since $G(P)$ reaches $G(P)/s_1$ under $V$, there exists a trace $u\in V$ such
that $G(P)/u = G(P)/s_1 = \ol n$. Since $s = s_1.e.s_2$ is a trace of $P$ and
$G(P)/s_1 = \ol n$, then $(\ol n,e)$ is in the domain of $t$. So, $ G(P)/u.e
= G(P)/s_1.e = n$ is a well-defined node of $N$ not contained in $t(\ii
n,V)$. Since $u.e\in V\cup V.\Sigma$, $G(P)$ reaches at least the additional
node $n$ under $V\cup V.\Sigma$. This completes the proof. \xbox
\end{proof}

% =========================================================================
\subsubsection*{Graph Products}
\label{sec:GP}

For proving our main theorems, it is necessary to consider the \emph{product}
of normalised transition graphs. We need this only for the investigation of
corresponding traces in reference processes and processes for SUTs. So, the
labelling of nodes with maximal refusals or minimal acceptances are
disregarded in the product construction. We consider two normalised
transition graphs
\[
G_i = ( N_i, \ii n_i, \Sigma, t_i : N_i\times\Sigma \pfun N_i, r_i : N_i \fun \mathbb{P}\mathbb{P}(\Sigma)),\qquad i = 1,2,
\]
over the same alphabet $\Sigma$. Their product is defined by
%
\begin{eqnarray}
G_1\times G_2 & = & (N_1\times N_2,(\ii n_1,\ii n_2), t:(N_1\times N_2)\times\Sigma\pfun (N_1\times N_2))
\\
\dom~t & = & \{ ((n_1,n_2),e)\in (N_1\times N_2)\times\Sigma~|   \nonumber
\\ & & \tabc
(n_1,e)\in\dom~t_1\wedge
(n_2,e) \in\dom~t_2    \}
\\
t((n_1,n_2),e) & = & (t_1(n_1,e),t_2(n_2,e))\ \text{for $((n_1,n_2),e)\in\dom~t$}
\end{eqnarray}
%
The following lemma is used in the proof of our main theorem.
%
\begin{lemma}\label{lemma:reachproduc}
If $G_1$ has $p$ states and $G_2$ has $q$ states, then every reachable state
$(n_1,n_2)$ of the product graph $G_1\times G_2$ can be reached by a trace
%%%$s\in\trc(G_1)\cap \trc(G_2)$  we don't need this
of maximal length $(pq-1)$.
\end{lemma}
\begin{proof}
The product graph $G_1\times G_2$ has at most $pq$ states. The empty trace $\varepsilon$
reaches its initial state $(\ii n_1,\ii n_2)$. Applying Lemma~\ref{lemma:extendV}
$(pq-1)$ times with $V=\{\varepsilon \}$ implies that $G_1\times G_2$ reaches
all of its reachable states (there are at most $pq$ of them) under
$V' = V \cup V.\Sigma\cup\dots \cup V.\Sigma^{(pq-1)}$. The maximal length of traces in
$V'$ is $(pq-1)$.
\xbox
\end{proof}
%
This concludes our presentation of CSP and of some results regarding its
semantics and normalised transition graphs that are used in the next section.

@@

In preparation for complexity results to be presented in
Section~\ref{sec:complexity}, we consider how many minimal hitting sets can
maximally exist for a given collection of minimal acceptances. To this end,
the following definitions and results are useful.

>>>>

We present CSP~(Section~\ref{section:csp}) and the concept of minimal hitting
sets~(Section~\ref{sec:hit}), which is central to our notion of test
for failures refinement. To study complexity, we also introduce the concept
of Sperner families~(Section~\ref{sec:sperner}).

@@

\paragraph{Communicating Sequential Processes (CSP)} This is a process algebra supporting
system development by refinement. Using CSP, we model both systems and their
components using processes. They are characterised by their patterns of
interactions, modelled by synchronous, instantaneous, and atomic events.

Throughout this paper, the alphabet of the processes, that is, the set of
events in scope, is denoted by $\Sigma$ and supposed to be finite.
The FDR tool~\cite{fdr} supports model checking and semantic analyses of
finite-state CSP processes.

@@

Two forms of choice support branching behaviour. An external choice $P
\extchoice Q$ between processes $P$ and $Q$ offers to the environment the
initial events of $P$ and $Q$. Once a synchronisation takes place, the
process that has offered the event that has occurred is chosen and the other
is discarded. In an internal choice $P \intchoice Q$, the environment does
not have an opportunity to interfere:~the choice is made by the process.
%
\begin{example}\label{example:CSP}
  We consider the processes $P$, $Q$, and $R$ defined below. $P$ is initially
  ready to engage in the event $a$, and then makes an internal choice to
  behave like either $Q$ or $R$.
  \begin{eqnarray*}
  P & = & a \then (Q\intchoice R)
  \\
  Q & = & a \then P \extchoice c \then P
  \\
  R & = & b \then P \extchoice c \then R
  \end{eqnarray*}
  $Q$, for instance, offers to the environment the choice to engage in $a$
  again or $c$. In both cases, afterwards, we have a recursion back to $P$.
  In $R$, if $b$ is chosen, we also have a recursion back to $P$. If $c$ is
  chosen, the recursion is to $R$.
  \xbox
\end{example}
%
Iterated forms $\Extchoice i: I @ P(i)$ and $\Intchoice i: I @ P(i)$ of the
external and internal choice operators define a choice over a collection of
processes $P(i)$. If the index set $I$ is empty, the external choice is the
process $\Stop$, which deadlocks:~does not engage into any event or
terminate. For an external choice $\Extchoice e: A @ e \then P(e), A\subseteq
\Sigma$ over a set $A$ of events, we use the abbreviation $e:A \then P(e)$.
An iterated internal choice is not defined for an empty index set.

There are several parallelism operators. A widely used form of parallelism $P
\parallel[cs] Q$ defines a process in which the behaviour is characterised by
those of $P$ and $Q$ in parallel, synchronising on the events in the set
$cs$. Other forms of parallelism available in CSP can be defined using this
parallelism operator.

@@

Semantic definitions specify, for each operator, how the traces or failures
of the resulting process can be calculated from those of each operand. For
example, for internal choice, $\fails(P\intchoice Q) =
\fails(P)\cup\fails(Q)$; see~\cite[p.~210]{Roscoe:1997:TPC:550448} for a
comprehensive list of these definitions covering $\trc(P)$ and $\fails(P)$.

Using the notation $P/s$ to denote the behaviour ot the process $P$ after
having engaged into the events in the trace $s$, the set $\refs(P/s) \defs
\{~X | (s,X) \in \fails(P)~\}$ contains the  refusals of $P$ after $s$.
Refusals are subset-closed~\cite{Hoare:1985:CSP:3921,Roscoe2010}: if $(s,X)$
is a failure of $P$ and $Y\subseteq X$, then $(s,Y)\in\fails(P)$ and
$Y\in\refs(P/s)$ follows.

For divergence-free processes, failures refinement, $P \lessdet_F Q$, is
defined by $\fails(Q)\subseteq\fails(P)$. Since refusals are subset-closed,
$P \lessdet_F Q$ implies $(s,\varnothing)\in\fails(P)$ for all traces
$s\in\trc(Q)$. So, for divergence-free processes, failures refinement implies
trace refinement. Therefore, using the conformance relation $conf$ below
%
\begin{equation}\label{eq:conf}
  Q\ conf\ P \defs \forall s\in \trc(P) \cap \trc(Q): \refs(Q/s)
  \subseteq \refs(P/s),
\end{equation}
%
failures refinement can be expressed by $\lessdet_T$ and $conf$ as proven
in~\cite{DBLP:conf/icfem/CavalcantiG07}.
%
\begin{equation}%\label{thm:fref-tref-conf}
\label{eq:failconf}
(P \lessdet_F Q) \iff (P \lessdet_T Q \land Q\ conf\ P)
\end{equation}
%
For finite processes, since refusals are subset-closed, $\refs(P/s)$ can be
constructed from the set of \emph{maximal refusals}.
%
\begin{equation}
\maxrefs(P/s) = \{ R \in\refs(P/s)~|~\forall R'\in \refs(P/s) - \{ R\}: R \not\subseteq R'\}
\end{equation}
%
Conversely, with the maximal refusals $\maxrefs(P/s)$ at hand, we can
reconstruct the refusals in the set $\refs(P/s)$. % as described by
%
\begin{equation}\label{eq:refMaxRef}
\refs(P/s) = \{ R'\in 2^\Sigma~|~\exists R\in \maxrefs(P/s): R'\subseteq R \}.
\end{equation}
%
Deterministic process states $P/s$ have exactly the one maximal refusal
defined by $\Sigma-[P/s]^0$, where $[P/s]^0$ denotes the \emph{initials} of
$P/s$, that is, the events that $P/s$ may engage into. The maximal refusals
in combination with the initials of a process express its degree of
nondeterminism as illustrated by the next example.
%
\begin{example}
\label{ex:nondetdegree} $P = (\Stop \intchoice Q)$ has maximal refusals
$\maxrefs(P) = \{ \Sigma \}$, because $\Stop$ refuses to engage in any event,
and this is carried over to $P$ by the internal choice. However, $P$ is
distinguished from $\Stop$ by its initials, which are defined by $[P]^0 = [
\Stop \intchoice Q]^0 = [Q]^0$. So $P$ may engage nondeterministically in any
initial event of $Q$, but also refuse everything, due to internal selection
of $\Stop$. Assuming an alphabet $\Sigma = \{a,b,c,d\}$, the process
%
$$Q = (e:\{a,b\} \then \Stop) \intchoice (e:\{c,d\} \then \Stop)$$
%
has maximal refusals $\maxrefs(Q) = \{ \{c,d\}, \{a,b\} \}$ and initials
$[Q]^0=\Sigma$. In contrast to $P$, nondeterminism is reflected here by
two maximal refusals. \xbox
\end{example}

% =========================================================================
\paragraph{Normalised Transition Graphs for CSP Processes} \label{sec:ntg}

As shown in~\cite{Roscoe:1994:chapter}, the failures semantics of any
finite-state CSP process $P$ can be represented by a \emph{normalised
transition graph} $G(P)$ defined by a tuple
$$
G(P) = ( N, \ii n, \Sigma, t : N\times\Sigma \pfun N, r : N \fun \mathbb{P}\mathbb{P}(\Sigma)),
$$
with nodes $N$, initial node $\ii n\in N$, and process alphabet $\Sigma$. The
partial \emph{transition function} $t$ maps a node $n$ and an event
$e\in\Sigma$ to its successor node $t(n,e)$. If $(n,e)$ is in the domain of
$t$, then there is a transition, that is, an outgoing edge, from $n$ with
label $e$, leading to node $t(n,e)$. Normalisation of $G(P)$ is reflected by
the fact that $t$ is a function.

The graph construction in~\cite{Roscoe:1994:chapter} implies that all nodes
$n$ in $N$ are reachable by sequences of edges labelled by $e_1\dots e_k$ and
connecting states $\ii n,n_1,\dots,n_{k-1},n$, such that
\[
n_1 = t(\ii n,e_1), \quad n_i = t(n_{i-1},e_i),\ i = 2,\dots,k-1,\quad
n= t(n_{k-1},e_k).
\]
%
By construction, $s\in\Sigma^*$ is a trace of $P$, if, and only if, there is
a path through $G(P)$ starting  at $\ii n$ whose edge labels coincide with
the events in $s$ in the order they appear. In analogy to $\trc(P)$, we use
the notation $\trc(G(P))$ for the set of finite, initialised paths through
$G(P)$, each path represented by its finite sequence of edge labels. We note
that $\trc(P) = \trc(G(P))$. Since $G(P)$ is normalised, there is a unique
node reached by following the events from $s$ one by one, starting in $\ii
n$. Therefore, $G(P)/s$  is also well defined.

By $[n]^0$ we denote the \emph{initials} of $n$:~the set of events occurring
as labels in any outgoing transitions.
$$
[n]^0 = \{ e\in\Sigma~|~(n,e)\in\dom~t \}
$$
The graph construction from~\cite{Roscoe:1994:chapter} guarantees that $[G(P)/s]^0 = [P/s]^0$ for all
traces $s$ of $P$.

@@

% =========================================================================
\paragraph{Acceptances} \label{sec:accs}

When investigating tests for failures refinement, the notion of
\emph{acceptances}, which is dual to refusals, is useful. While the
original introduction of acceptances presented
in~\cite[pp.~75]{Hennessy:1988:ATP:50497} was independent of refusals, we use
the definition from~\cite[pp.~278]{Roscoe:1997:TPC:550448}.   A \emph{minimal
acceptance} of a CSP process state $P/s$ is the complement of a maximal
refusal of the same state. The set of minimal acceptances of $P/s$ is denoted
by $\minaccs(P/s)$ and formally defined as
%
\begin{equation}\label{eq:accref}
\minaccs(P/s) = \{ \Sigma - R~|~R\in \maxrefs(P/s)  \}
\end{equation}
%
With this definition, a (not necessarily minimal) \emph{acceptance} of $P/s$ is a superset of some minimal acceptance and a subset of the initials $[P/s]^0$. Denoting the acceptances of $P/s$ by $\accs(P/s)$, this leads to the formal definition
\begin{equation}
\label{eq:accrefall}
\accs(P/s)  = \{ B\subseteq [P/s]^0~|~\exists A\in\minaccs(P/s):  A \subseteq B   \}
\end{equation}
%
Acceptances have the following intuitive interpretation.  If the behaviour of
$P/s$ is deterministic, its only acceptance equals $[P/s]^0$, because $P/s$
never refuses any of the events in this set. If $P/s$ is nondeterministic, it
internally chooses one of its \emph{minimal acceptance sets} $A$ and never
refuses any event in $A$, while {\it possibly} refusing the events from the
set $[P/s]^0 - A$ and {\it always} refusing those in the set $\Sigma -
[P/s]^0$.

Exploiting (\ref{eq:accref}), the nodes of a normalised transition graph can
alternatively be labelled with minimal acceptances; this captured the same
information conveyed by maximal refusals. Since process states $P/s$ are
equivalently expressed by states $G(P)/s$ of $P$'s normalised transition
graph, we also write $\minaccs(G(P)/s)$ and note that (\ref{eq:refsAndR}) and
(\ref{eq:accref}) imply
\begin{equation}\label{eq:maxrefsminaccs}
\minaccs(G(P)/s) = \{ \Sigma - R | R\in r(G(P)/s) \} = \minaccs(P/s).
\end{equation}
%
Given any non-diverging, non-terminating, finite-state process $P$, it can be
re-con\-struc\-ted from its graph $G(P)$ with initial state $\ii
n$ and transition function $t$, using $P$'s normalised syntactic
representation~\cite[pp.~277]{Roscoe:1997:TPC:550448} specified as follows.
\begin{eqnarray*}
\text{normalised}(P) & = & P_N(\ii n)
\\
P_N(n) & = & \Intchoice_{A\in\minaccs(n)\cup \{ [n]^0 \} } e:A\then P_N(t(n,e))
\end{eqnarray*}
With this definition, it is established that $P$ is semantically equivalent
to $\text{normalised}(P)$ in the failures semantics.

% .....................................................................................
 \begin{figure}[t]
   %%\hspace*{-40mm}
   \begin{center}
     \includegraphics[trim=0cm 1cm 0cm 1cm,clip,scale=.6,%width=\textwidth
                              ]{p.pdf}
   \end{center}
   %%\vspace*{-10mm}
   \caption{Normalised transition graph of CSP process $P$ from Example~\ref{ex:a}.}
   \label{fig:tga}
 \end{figure}
% .......................................................................................

@@

node~3 corresponds to $R/c = R$, and so it is labelled by $R$'s minimal
acceptance $\{b,c\}$. \xbox

@@

\noindent%
Summarising, refinement between finite-state CSP processes $P, Q$ can be
expressed using their normalised graphs

@@

%  \\
%  \label{eq:failconf}
%  P \lessdet_F Q & \Leftrightarrow & P \lessdet_T Q \wedge Q\ conf\ P

@@

To prove (\ref{eq:trcrefa}), we recall that $P\lessdet_T Q$ is defined as
$\trc(Q)\subseteq \trc(P)$ and, moreover, $\trc(P)=\trc(G(P))$ and
$\trc(Q)=\trc(G(Q))$. %Result (\ref{eq:failconf}) has been shown
%in~\cite{DBLP:conf/icfem/CavalcantiG07}.

@@

@@

  \\ %

@@

% =========================================================================
\paragraph{Reachability Under Sets of Traces} \label{sec:V} Given a finite-state CSP
process $P$ and its normalised transition graph $G(P)$,
%\[
%G(P) = ( N, \ii n, \Sigma, t : N\times\Sigma \pfun N, r : N \fun \mathbb{P}\mathbb{P}(\Sigma)),
%\]
we suppose that $V\subseteq\Sigma^*$ is a prefix-closed set  of sequences of
events. By $t(\ii n,V)$ we denote the set
\[
t(\ii n,V) = \{ n\in N~|~\exists s\in V: s\in\trc(P)\wedge G(P)/s = n \}
\]
of nodes in $N$ that are reachable in $G(P)$ by applying traces of $V$. The
lemma below specifies a construction method for such sets $V$ reaching {\it
every} node of $N$.

\begin{lemma}
\label{lemma:extendV} Let $P$ be a CSP process with normalised transition
graph $G(P)$. Let $V\subseteq\Sigma^*$ be a finite prefix-closed set of
sequences of events. Suppose that  $G(P)$ reaches $k < |N|$ nodes under $V$,
that is, $|t(\ii n,V)| = k$. Let $V.\Sigma$ denote the set of all sequences
from $V$, extended by any event of $\Sigma$.  Then $G(P)$ reaches at least
$(k+1)$ nodes under $V\cup V.\Sigma$.
\end{lemma}
\begin{proof}
Suppose that $n'\in (N - t(\ii n,V))$.  Since all nodes in $N$ are reachable,
there exists a trace $s$ such that $G(P)/s = n'$. Decompose $s = s_1.e.s_2$
with $s_i\in\Sigma^*, e\in\Sigma$, such that $G(P)/s_1 \in t(\ii n,V)$ and
$G(P)/s_1.e \not\in t(\ii n,V)$. Such a decomposition always exists, because
$V$ is prefix-closed and therefore contains the empty trace $\varepsilon$.
Note, however, that it is not necessarily the case that $s_1\in V$. Since
$G(P)$ reaches $G(P)/s_1$ under $V$, there exists a trace $u\in V$ such that
$G(P)/u = G(P)/s_1 = \ol n$. Since $s = s_1.e.s_2$ is a trace of $P$ and
$G(P)/s_1 = \ol n$, then $(\ol n,e)$ is in the domain of $t$. So, $ G(P)/u.e
= G(P)/s_1.e = n$ is a well-defined node of $N$ not contained in $t(\ii
n,V)$. Since $u.e\in V\cup V.\Sigma$, $G(P)$ reaches at least the additional
node $n$ under $V\cup V.\Sigma$. This completes the proof. \xbox
\end{proof}

% =========================================================================
\paragraph{Graph Products} \label{sec:GP} For proving our main theorems, it is
necessary to consider the \emph{product} of normalised transition graphs. We
need this only for the investigation of corresponding traces in reference
processes and processes for SUTs. So, the labelling of nodes with maximal
refusals or minimal acceptances are disregarded in the product construction.
We consider two normalised transition graphs
\[
G_i = ( N_i, \ii n_i, \Sigma, t_i : N_i\times\Sigma \pfun N_i, r_i : N_i \fun \mathbb{P}\mathbb{P}(\Sigma)),\qquad i = 1,2,
\]
over the same alphabet $\Sigma$. Their product is defined by
%
\begin{eqnarray}
G_1\times G_2 & = & (N_1\times N_2,(\ii n_1,\ii n_2), t:(N_1\times N_2)\times\Sigma\pfun (N_1\times N_2))
\\
\dom~t & = & \{ ((n_1,n_2),e)\in (N_1\times N_2)\times\Sigma | (n_1,e)\in\dom~t_1\wedge (n_2,e) \in\dom~t_2 \}
\\
t((n_1,n_2),e) & = & (t_1(n_1,e),t_2(n_2,e))\ \text{for $((n_1,n_2),e)\in\dom~t$}
\end{eqnarray}
%
The following lemma is used in the proof of our main theorem.
%
\begin{lemma}\label{lemma:reachproduc}
If $G_1$ has $p$ states and $G_2$ has $q$ states, then every reachable state
$(n_1,n_2)$ of the product graph $G_1\times G_2$ can be reached by a trace
%%%$s\in\trc(G_1)\cap \trc(G_2)$  we don't need this
of maximal length $(pq-1)$.
\end{lemma}
\begin{proof}
The product graph $G_1\times G_2$ has at most $pq$ states. The empty trace $\varepsilon$
reaches its initial state $(\ii n_1,\ii n_2)$. Applying Lemma~\ref{lemma:extendV}
$(pq-1)$ times with $V=\{\varepsilon \}$ implies that $G_1\times G_2$ reaches
all of its reachable states (there are at most $pq$ of them) under
$V' = V \cup V.\Sigma\cup\dots \cup V.\Sigma^{(pq-1)}$. The maximal length of traces in
$V'$ is $(pq-1)$.
\xbox
\end{proof}
%
This concludes our presentation of CSP and of results regarding its
semantics that are used in the next section.

@@

In preparation for complexity results presented in
Section~\ref{sec:complexity}, we consider how many minimal hitting sets can
maximally exist for a collection of minimal acceptances. To this end, the
following definitions and results are useful.
